Visualizing Contingency Tables

### 2006-05-01 trunk/vcd/paper/hsc.Rnw
next revision ...
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@525 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
Chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

@@

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression. %FIXME% mention this?
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot.

@@

In a first step, we create a mosaic plot for an exploratory view on the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used as the last split via highlighting. 
Furthermore, we expect age and education to have an influence on both the memory
and the attitude, hence they should be used first for splitting as stratifying
variables. In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people have more than 40 years of
age, most of which of only completed elementary school (much more than 
in the other age groups). Furthermore, it can be seen that the proportion
of people with memories of punishments increases with age and that
the approval rate decreases with education. As for the question whether
attitude depends on memory, the plot suggests quite clearly (and somewhat
surprisingly) that for those people with elementary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without: 
experienced violence seems to engender violence again.
For the other education groups, the picture is less clear: some cells 
indicate the same association while others do not.

@@

In summary, both age and education clearly matter here.
But is there really any significant influence of memory on the attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
Are memory and attitude conditionally independent, given age and education?
To focus on this question, we will take another model-based view on the data.
To reduce the complexity of the information to the question of interest, we employ
partial mosaic plots in a trellis layout (conditioning on age and education) with
residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
This visualization brings out much more clearly that the increased approval rate
among people with memories of punishments is present mainly in the elementary education
column and, to a lesser degree, in the age 40+ row. Moreover, this association is
only significant in two cells (the two older age groups with elementary education)
where fewer people without memories do not approve punishment than expected under
(conditional) independence.

>>>>

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
Chapter III-12 in this book provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter III-12.

@@

Another approach is to visualize (the deviations from) a particular fitted model 
with similar ideas as in Section~\ref{sec:extensions}. Thus, mosaic displays can
be extended from purely explorative views to model-based views on the data by
residual-based shadings and association plots can directly visualize deviations
from a given model. Typically, the models considered are log-linear models, however
residual-based shadings are also conceivable for other types of models such as logistic
regression. %FIXME% mention this?
In our example, from the exploratory analysis, we already suspect that
admission and gender are independent, given department, except for
department A. 
Therefore, a sensible way of representing the table is to 
nest admission into department instead of gender to get a stratified view.
The corresponding mosaic and association plots are shown
in Figures~\ref{fig:ucbmosaic} and \ref{fig:ucbassoc}. The shading visualizes the 
model of admission and gender being conditionally independent, given department. Clearly,
there is no gender bias in the departments except for department A: here, 
significantly \textit{more} female students are admitted then expected under independence.
Mosaic displays are already recognized as an excellent
visualization tool for log-linear models 
\cite{vcd:Friendly:1999,vcd:Theus+Lauer:1999,vcd:Hofmann:2001}.
Using the flat table representation, the association plot similarly extends from
the simple two-way case to a structured residual plot usable for 
the diagnostics of log-linear models. Especially if a
mosaic display contains very small or empty cells, the corresponding association plot
might provide an easier way to detect deviation patterns than the
residual-based shading added to the mosaic plot. Yet another approach
is to plot a mosaic plot of the \emph{expected} frequencies and to highlight
the (relative) residuals in the cells, using different colors for
positive and negative residuals \cite{vcd:Theus+Lauer:1999}.

@@

In a first step, we create a mosaic plot for an exploratory view on the data:
Attitude towards punishment clearly is the dependent variable 
here and should be used as the last split via highlighting. 
Furthermore, we expect age and education to have an influence on both the memory
and the attitude, hence they should be used first for splitting as stratifying
variables. In fact, the question is whether memory has some influence on attitude, given 
age and education. We choose to create a three-way mosaic plot of the 
hypothesized explanatory variables (first splitting horizontically by age, 
then vertically by education, and finally horizontically again by memory), 
and to have attitude, the dependent variable, 
highlighted in the tiles (see Figure~\ref{fig:punishhigh}).
We can see that half of the people have more than 40 years of
age, most of which of only completed elementary school (much more than 
in the other age groups). Furthermore, it can be seen that the proportion
of people with memories of punishments increases with age and that
the approval rate decreases with education (this, however, is better shown in 
Figure~\ref{fig:punishcotab} described below). As for the question whether
attitude depends on memory, the plot suggests quite clearly (and somewhat
surprisingly) that for those people with elementary school education,
a higher proportion \emph{with} memories tends to
accept moderate punishment of children than those without: 
experienced violence seems to engender violence again.
For the other education groups, the picture is less clear: some cells 
indicate the same association while others do not.

@@

In summary, both age and education clearly matter here.
But is there really any significant influence of memory on the attitude towards 
punishment apart from that? Or, rephrased as a hypothesis:
Are memory and attitude conditionally independent, given age and education?
To focus on this question, we will take another model-based view on the data
that avoids the distracting influence of the marginal variables (age
and education): We employ
partial mosaic plots in a trellis layout (conditioning on age and education) with
residual-based shading and data-driven cut-offs (see Figure~\ref{fig:punishcotab}).
This visualization brings out much more clearly that the increased approval rate
among people with memories of punishments is present mainly in the elementary education
column and, to a lesser degree, in the age 40+ row. Using, e.g., a 
permutation test similar to the one discussed in Section \ref{sec:test}, 
model statistics reveal that the hypothesis is rejected at a significance level of 
1\%, but the shading clearly shows that this association is
only significant in two cells (the two older age groups with elementary education)
where fewer people without memories do not approve punishment than expected under
(conditional) independence. In the group of people with age between
25--39, the lighter shading indicates that here,
the association is only significant at the 10\% level.
